---
title: "PathoFMPred: building and using portable pathology models"
author: "PathoFMPred authors"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{PathoFMPred: building and using portable pathology models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Scope and distribution

PathoFMPred stores many cancer-specific PLS regression and PLS-LDA fits in one
portable object for each pathology representation. The object contains fitted
parameters, feature order, development metrics, and provenance. It does not
contain the original feature or outcome rows.

The public package contains only two small Giga-SSL models as executable
examples. Full Giga-SSL and Prov-GigaPath collections are optional downloads
that occur only when a user calls `fetch_pathofmpred_models()`. No model is
downloaded during installation, package loading, checking, or vignette
construction.

TITAN-derived fitted parameters are not distributed in the public repository.
Users may build a TITAN object locally after separately obtaining the TITAN
features and accepting the upstream access terms. Do not redistribute the
resulting object unless the TITAN rights holder has provided permission.

The PathoFMPred source code is MIT licensed. Fitted objects, registry data and
third-party inputs are not covered by that source-code grant. The complete
asset-specific terms and required attributions are recorded in
`inst/licenses/MODEL_ACCESS.md`.

All supplied performance estimates are internal research estimates. They do
not establish external validity, assay equivalence, or clinical utility.

PathoFMPred enforces the fastPLS version recorded in each fitted object. When
the R installation records the optional `RemoteSha` field, the exact Git
revision is also enforced. Base installation from an immutable source archive
does not populate `RemoteSha`; PathoFMPred then issues one provenance warning
per session and proceeds only if the recorded package version matches.

# Install optional public model collections

```{r fetch, eval=FALSE, purl=FALSE}
library(PathoFMPred)
fetch_pathofmpred_models(c("GigaSSL", "ProvGigaPath"))
available_foundation_models()
```

An institutional mirror can be selected without changing package code:

```{r mirror, eval=FALSE, purl=FALSE}
fetch_pathofmpred_models(
  "GigaSSL",
  base_url = "https://github.com/tkcaccia/PathoFMPred/releases/download/models-v2"
)
```

# Build a PathoFMPred object from a new representation

The first three arguments are mandatory. Both tables must contain the column
named by `id_column`. The outcome table must contain unique IDs. The feature
table may contain repeated IDs, such as several slides from one patient.
Repeated rows are pooled by patient mean by default or by patient median when
`aggregation = "median"` is selected. More than 100 unique IDs must match.

```{r build-generic, eval=FALSE, purl=FALSE}
features <- read.csv("foundation_model_features.csv", check.names = FALSE)
outcomes <- read.csv("patient_outcomes.csv", check.names = FALSE)

custom_object <- create_pathofmpred_object(
  feature_table = features,
  outcome_table = outcomes,
  id_column = "patient_id",
  foundation_model = "my_future_representation",
  aggregation = "mean",
  output_file = "pathofmpred_my_future_representation.rds"
)
```

The fitted collection can then be applied directly without adding it to the
package's bundled-model registry. Repeated rows in a new feature table are
pooled with the aggregation rule stored in the object.

```{r predict-generic, eval=FALSE, purl=FALSE}
new_features <- read.csv("new_foundation_model_features.csv",
                         check.names = FALSE)

custom_prediction <- predict_pathofmpred_object(
  object = custom_object,
  features = new_features,
  id_column = "patient_id"
)
custom_prediction[, c("patient_id", "endpoint", "outcome_type",
                      "prediction", "predicted_class_label", "lda_score")]
```

For a binary endpoint, `predicted_class_label` uses the negative and positive
labels recorded during fitting. `lda_score` is an uncalibrated discriminant
score and is not a probability.

The function reports feature-only IDs, outcome-only IDs, and missingness for
every endpoint in `custom_object$data_audit`. Numeric outcomes with two observed
levels are treated as binary by default. Other numeric outcomes are treated as
continuous. Types can be fixed explicitly with a named vector. Binary outcomes
encoded as anything other than `0`/`1` or `FALSE`/`TRUE` require an explicit
event label through `positive_class`.

```{r settings, eval=FALSE, purl=FALSE}
settings <- pathofmpred_control(
  components = 1:20,
  outer_folds = 5,
  inner_folds = 5,
  repeats = 5,
  method = "plssvd",
  binary_selection = "AUROC",
  seed = 2026
)

custom_object <- create_pathofmpred_object(
  features, outcomes, "patient_id",
  outcome_types = c(APC = "binary", Leukocyte_Fraction = "continuous"),
  positive_class = c(APC = "mutated"),
  control = settings
)
```

`rsvd_oversample` and `rsvd_power` are `NULL` by default. Therefore,
PathoFMPred uses the installed fastPLS defaults. They can be overridden through
`pathofmpred_control()` when a documented analysis requires different values.

# Build the TITAN object locally

The official TITAN TCGA slide representations are provided by the TITAN
authors in the gated
[MahmoodLab/TITAN repository](https://huggingface.co/MahmoodLab/TITAN),
including `TCGA_TITAN_features.pkl`. Access must be requested from the upstream
repository. The project conversion script is
`tools/convert_titan_features.py` in the companion analysis repository. It
creates a CSV whose rows retain TCGA slide identifiers and whose 768 feature
columns are named consistently.

Outcome tables must be assembled independently. The companion analysis
repository documents and converts the following sources:

* MC3 mutation calls from Ellrott et al. and cancer-driver annotations from
  Bailey et al.;
* PanImmune variables from Thorsson et al., including RNA-deconvolved cell
  fractions, transcriptomic signatures, methylation-derived leukocyte content,
  immune-repertoire quantities, in-silico neoantigens, and genomic-context
  variables;
* oncogenic pathway alterations from Sanchez-Vega et al.;
* aneuploidy and genome-doubling variables from Taylor et al.;
* gene fusions from Gao et al.;
* MSI fields from the cBioPortal TCGA PanCancer Atlas studies; and
* RNA pathway scores computed from TCGA expression matrices, including cancer
  hallmarks, glutathione, and vitamin B6 pathways.

The exact expected filenames and local configuration are recorded in
[`config/paths.example.R`](https://github.com/tkcaccia/titan-prediction/blob/main/config/paths.example.R).
The acquisition and conversion workflow is documented in the companion
[`README`](https://github.com/tkcaccia/titan-prediction). Molecular sources do
not all link to histology at the same level. Some provide a TCGA sample barcode,
whereas the PanImmune table links at participant level. A matching sample
barcode does not prove that the slide and molecular assay came from the same
block, portion, analyte, or tumour region.

After conversion, a local TITAN collection can be fitted as follows. This code
is intentionally not executed while the vignette is built.

```{r build-titan, eval=FALSE, purl=FALSE}
titan_features <- read.csv(
  "TCGA_TITAN_features.csv", check.names = FALSE
)
endpoint_table <- read.csv(
  "pathofmpred_patient_endpoints.csv", check.names = FALSE
)

pathofmpred_titan <- create_pathofmpred_object(
  feature_table = titan_features,
  outcome_table = endpoint_table,
  id_column = "patient_id",
  foundation_model = "TITAN",
  aggregation = "mean",
  control = pathofmpred_control(
    components = 1:20,
    outer_folds = 5,
    inner_folds = 5,
    repeats = 5,
    method = "plssvd",
    binary_selection = "AUROC"
  ),
  output_file = "pathofmpred_titan.rds"
)

titan_prediction <- predict_pathofmpred_object(
  object = pathofmpred_titan,
  features = read.csv("new_TITAN_features.csv", check.names = FALSE),
  id_column = "patient_id"
)
```

Keep `pathofmpred_titan.rds` local or in the access-controlled
`tkcaccia/PathoFMPred-private` repository. It must not be added to the public
`tkcaccia/PathoFMPred` repository under the current release policy.

# Use the bundled Giga-SSL example

The minimal public fixture contains COAD APC mutation and MANTIS-score models.
It demonstrates the interface but is not a complete atlas.

```{r example, eval=FALSE, purl=FALSE}
models <- available_pathofm_models("GigaSSL", cancer = "COAD")
models[, c("family", "endpoint", "outcome_type", "n")]

x <- example_pathofm_features("COAD", foundation_model = "GigaSSL")
prediction <- predict_pathofm(
  cancer = "COAD",
  features = x,
  foundation_model = "GigaSSL",
  patient_id = "software_example"
)
prediction[, c("endpoint", "outcome_type", "prediction",
               "predicted_class", "lda_score")]
```

Binary scores are uncalibrated discriminant scores, not probabilities. Any
reference rank is a rank within TCGA repeated out-of-fold predictions, not a
probability or clinical-risk estimate.
